Inferring community characteristics in labelled networks

### 2021-05-26 nips/derivations.tex
More hyp testing

We are given samples $\{\theta^{(t)}\}_{t \in \Tcal_\theta} \sim p(\theta | A, X)$ and wish to determine the statistical significance of the weights. We adopt matrix notation for simplicity by representing $\theta$ with the matrix $B \times D$ matrix of feature weights $W$. The question is to determine the significance of a particular feature $d$ by examining the value of $W_{id}$ for all the values $i \in [B]$. We know that the posterior is proportional to the prior multiplied by the likelihood (assuming that the feature matrix $X$ is already given):
%
\begin{equation}
	p(\theta|A, X) \propto p(\theta) \cdot p(A | \theta, X)
\end{equation}
%
The prior term can be evaluated but the likelihood is intractable. The closest we can get is through a Monte-Carlo integration over $b$:
%
\begin{align}
	p(A | \theta, X) &= \sum_{b \in [B]^N} p(A, b | \theta, X) \nonumber \\
	&= \sum_{b \in [B]^N} p(A | b, \theta, X) \cdot p(b | \theta, X) \nonumber \\
	&= \sum_{b \in [B]^N} p(A | b) \cdot p(b | \theta, X) \nonumber \\
	&\approx \sum_{i} p\left( A | b^{(i)} \right) \quad \textrm{with} \quad b^{(i)} \sim p(b| \theta, X)
	\label{eqn:mc-likelihood}
\end{align}
%
This could be implemented for a single value of $\theta$ but such a form cannot be used to characterise the overall form of the posterior. Nevertheless, the form in equation \ref{eqn:mc-likelihood} still does highlight something interesting. The likelihood is peaked around areas of $\theta$ that generate a partition $b$ that is highly likely in the SBM sense -- high $p(A|b)$. This provides the motivation for using the Laplace approximation for modelling the posterior $p(\theta | A, X) \approx p(\theta; \mu, \Sigma)$. Indeed, the Laplace approximation is often used for modelling the posterior in logistic classification \cite{laplace}. As a simplification we assume that $\Sigma$ is diagonal, therefore each element of $\theta$ is independent. This assumption does not hold in general but it motivates the derivation of the dimensionality reduction we construct by analogy with hypothesis testing. Assuming, $\Sigma$ is diagonal then the posterior for each element of the weight matrix $W$ can be approximated by:
%
\begin{equation}
	p(W_{ij}|A, X) \approx \Gaussian(W_{ij}|\hat{\mu}_{ij}, \hat{\sigma}_{ij}^2)
\end{equation}
%
Where we have used the set of samples for $W$ drawn according to the exact posterior, to calculate unbiased estimates for the mean and standard deviation:
%
\begin{equation}
	\hat{\mu}_{ij} \coloneqq \frac{1}{|\Tcal_\theta|} \sum_{t \in \Tcal_\theta} W_{ij}^{(t)} \qquad \textrm{and} \qquad
	\hat{\sigma}_{ij}^2 \coloneqq \frac{1}{|\Tcal_\theta|} \sum_{t \in \Tcal_\theta} \left( W_{ij}^{(t)} - \hat{\mu}_{ij} \right)^2
\end{equation}
%
The probability

>>>>

We are given samples $\{\theta^{(t)}\}_{t \in \Tcal_\theta} \sim p(\theta | A, X)$ and wish to determine the statistical significance of the weights. We adopt matrix notation for simplicity by representing $\theta$ with the matrix $B \times D$ matrix of feature weights $W$. The question is to determine the significance of a particular feature $d$ by examining the value of $W_{id}$ for all the values $i \in [B]$. We know that the posterior is proportional to the prior multiplied by the likelihood (assuming that the feature matrix $X$ is already given):
%
\begin{equation}
	p(\theta|A, X) \propto p(\theta) \cdot p(A | \theta, X)
\end{equation}
%
The prior term can be evaluated but the likelihood is intractable. The closest we can get is through a Monte-Carlo integration over $b$:
%
\begin{align}
	p(A | \theta, X) &= \sum_{b \in [B]^N} p(A, b | \theta, X) \nonumber \\
	&= \sum_{b \in [B]^N} p(A | b, \theta, X) \cdot p(b | \theta, X) \nonumber \\
	&= \sum_{b \in [B]^N} p(A | b) \cdot p(b | \theta, X) \nonumber \\
	&\approx \sum_{i} p\left( A | b^{(i)} \right) \quad \textrm{with} \quad b^{(i)} \sim p(b| \theta, X)
	\label{eqn:mc-likelihood}
\end{align}
%
This could be implemented for a single value of $\theta$ but such a form cannot be used to characterise the overall form of the posterior. Nevertheless, the form in equation \ref{eqn:mc-likelihood} still does highlight something interesting. The likelihood is peaked around areas of $\theta$ that generate a partition $b$ that is highly likely in the SBM sense -- high $p(A|b)$. This provides the motivation for using the Laplace approximation for modelling the posterior $p(\theta | A, X) \approx p(\theta; \mu, \Sigma)$. Indeed, the Laplace approximation is often used for modelling the posterior in logistic classification \cite{laplace}. As a simplification we assume that $\Sigma$ is diagonal, therefore each element of $\theta$ is independent. This assumption does not hold in general but it motivates the derivation of the dimensionality reduction we construct by analogy with hypothesis testing. Assuming, $\Sigma$ is diagonal then the posterior for each element of the weight matrix $W$ can be approximated by:
%
\begin{equation}
	p(W_{ij}|A, X) \approx \Gaussian(W_{ij}|\hat{\mu}_{ij}, \hat{\sigma}_{ij}^2)
\end{equation}
%
Where we have used the set of samples for $W$ drawn according to the exact posterior, to calculate unbiased estimates for the mean and standard deviation:
%
\begin{equation}
	\hat{\mu}_{ij} \coloneqq \frac{1}{|\Tcal_\theta|} \sum_{t \in \Tcal_\theta} W_{ij}^{(t)} \qquad \textrm{and} \qquad
	\hat{\sigma}_{ij}^2 \coloneqq \frac{1}{|\Tcal_\theta|} \sum_{t \in \Tcal_\theta} \left( W_{ij}^{(t)} - \hat{\mu}_{ij} \right)^2
\end{equation}
%
This approximation is not exact but can show it is accurate empirically. Indeed, if we run the primary school experiment with hyper-parameters given in appendix \ref{appdx:hyperparams}, we can then plot histograms of the collected $W$-samples and compare these to the Laplace approximation. The results are given on figure \ref{fig:school-histogram}. Even though the Laplace approximation is not exact, it is remarkably reliable.

\begin{figure}[!h]
	\centering
	\begin{subfigure}[t]{0.3\linewidth}
		\centering
		\includegraphics[width=\linewidth]{school-sample-histogram-16.png}
		\caption{Weight for block 1, class 3B}
	\end{subfigure}
	\hfill
	\begin{subfigure}[t]{0.3\linewidth}
		\centering
		\includegraphics[width=\linewidth]{school-sample-histogram-37.png}
		\caption{Weight for block 3, class 4A}
	\end{subfigure}
	\hfill
	\begin{subfigure}[t]{0.3\linewidth}
		\centering
		\includegraphics[width=\linewidth]{school-sample-histogram-84.png}
		\caption{Weight for block 8, class 2B}
	\end{subfigure}

	\caption{Histograms of sampled weights for the primary school experiment \cite{schools}. Dotted line is the applied Laplace approximation.}
	\label{fig:school-histogram}
\end{figure}

Using the approximation, we can construct a test to determine whether a particular weight value has $|W_{ij}| > c$ with high probability. Assuming without loss of generality that
%
\begin{equation}
	p(|W_{ij}| > c) = p(W_{ij})
\end{equation}
